Evolution of Methodologies in Delivering Reliable Software in the European Scientific Arena

### 2018-05-28 paper-ieee-trans.tex

\IEEEPARstart{D}uring the last 15 years the provision of an European e--Infrastructure,
supporting an unified access to large--scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan--European research collaborations. One such example is the Worldwide
Large Hadron Collider (LHC) Computing Grid infrastructure, driven
by the particle physics scientific community, built to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
LHC particle accelerator. A significant number of innovation projects and
technology transfer activities -- from the European Commission (EC)-funded 5th, 6th and 7th Framework Programmes,
and recently from the Horizon 2020 programme \cite{h2020} -- have as their main goal to
develop software solutions that provide a seamless and transparent access of researchers to distributed
computing and data resources.

Major developments were carried out throughout the years focused on providing new
functionalities, partially neglecting the adoption of software engineering methodologies, in the early days.
The released software was less tested, resulting in a high rate of
rollbacks and patches applied to production systems. Thereby, the stability and
reliability of the e--Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.

@@

Analysing the EC--funded software projects over the last decade,
one can observe a continuous increase in the prominence and robustness of software
testing and validation procedures. The
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and
operating--system--level virtualization known as containerization) definitely motivated this trend. All of which
complemented by the emergence of automation and event--response tools.

@@

\begin{table*}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Features on software reliability in EC--projects}
\label{tab:feat}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{adjustbox}{max width=0.5\textwidth}
\begin{tabular}{llllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
\hline
\hline
Architecture Task Force&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection&&&&$\surd$&&$\surd$\\
Build/Testing Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&&$\surd$\\
Software Product Metrics&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$\\
%Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
Automatic Certification&&&$\surd$&&$\surd$&$\surd$\\
Auto--generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD)&&&&&&$\surd$\\
\hline
\hline
\end{tabular}
%\end{adjustbox}
\end{table*}

\subsection{Enabling Grids for E--sciencE}

The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
scientists and engineers from over 240 institutions in 45 countries aiming to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization (outside Europe) of the project, embracing worldwide research
institutions and user communities. The software to sustain the increasing
requirements coming from the diverse scientific communities needed developing a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15 thousand researchers and deployed in
over 250 institutions.

@@

The {\sl INDIGO--DataCloud} (INtegrating Distributed data Infrastructures for Global
explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this paper. At the time of writing
this paper, the project is succeeding in providing solutions to address the existing gaps in
Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructure administrators and scientific communities to exploit
Cloud computing benefits.

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper and thus
leveraged the lessons learned from the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
quality of the implemented solutions. All the above with the additional complexity of being framed into
large collaborative projects with an extensive list of partners.

In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition} provides guidelines to develop quality
software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing prompt reactions to detected malfunctions throughout the
process. A \textit{bottom--up adoption of DevOps principles} promotes automation to enable
the foregoing change--based development process and set up advanced delivery mechanisms.
The last key pillar is the \textit{two--factor validation} carried out over the project's pilot testbeds and
e--Infrastructure resource provider's early--adoption.

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=\textwidth]{images/fig2.png}
\caption{Comparison of the unit testing coverage values for the {\sl INDIGO--DataCloud} software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases.}
\label{fig:fig_unittest}
\end{figure*}

@@

\begin{itemize}
\item \textit{Code style}.
The ultimate goal for code style assessment is to improve the readability and reusability of the source code produced under the scope of the
project. Code style standards are enforced for every software component in the project stack. Community
most--adopted or \textit{de--facto standards} are initially recommended, however the development teams eventually
choose the set of guidelines to comply with.  Figure \ref{fig:fig_codestyle} highlights the code style standards and their popularity
in the {\sl INDIGO--DataCloud} software stack.

@@

 \textit{Unit testing} completes the source code testing evaluation by
focusing on the code's internal design. The SQA criteria set a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component between the two major {\sl INDIGO--DataCloud} releases, denoting an incremental
trend over the course of the project. It can be seen that 50\% of the components were over the
70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.

@@

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has
successfully passed through the testing methods described previously. It implies the human--based
revision of the proposed change to discuss its adequacy in terms of e.g. scope, objective fulfilment,
and documentation completeness. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
important when the SQA testing requirements are fully automated. Secure code reviews are also
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

\item \textit{Automated Deployment}.
To lower the barriers of software adoption, the SQA criteria require the automated
deployment of the products delivered as part of the catalogue. Automation in this context is
tackled using a configuration management tool. These tools allow to have an
\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation
-- self--documenting code -- as they sequentially guide the component's deployment process in multiple
platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \cite{indigo-galaxy}. Figure
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release and in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\subsection{DevOps practices}
\label{sec:devops}

The DevOps methods emphasize the SQA techniques to avoid infrastructure
disruption whenever new developments are deployed into production systems. The
following DevOps approaches have been progressively adopted throughout the project's
lifetime, being one of its major outcomes in terms of software reliability.

@@

Jenkins \cite{jenkins}, the event--response CI system, and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs.
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline, report their exit status back to GitHub preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects increases. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected in this phase.

Figure \ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the
{\sl INDIGO--DataCloud}'s first and second releases, of the required test types builds since the
implementation of the CI infrastructure. Automated functional testing coverage were not
available for all the software stack, thus the associated number of builds are fewer.

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the software as
they involve fewer changes and usually are more correlated. The software updates of
{\sl INDIGO--DataCloud} products, taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI--validated
software components are automatically delivered in online repositories without the need of
any human interaction but ultimately requiring expert supervision to validate the results.

@@

The \textit{Docker container pipeline} uses the automated deployment solution to
actually install and configure the software component in the container image. Once this is
done the recently created image is uploaded to the production DockerHub repository, tagged
with the label corresponding to the current {\sl INDIGO--DataCloud} release. Figure
\ref{fig:fig_CD} shows the complete workflow of the Docker container CD implementation,
automatically triggered by a change in the source code.

@@

Two applications from "Definition of Support to Research Communities" work package,
DisVis \cite{disvis} and PowerFit \cite{powerfit}, were
integrated into a similar CI/CD pipeline described in section \ref{sec:devops}.
Figure \ref{fig:fig_disvis} shows the pipeline for the Disvis application.

@@

Two pilot infrastructures are at the disposal of developers and scientific
communities involved in the project. The aim of these testbeds is to test the
level of integration between the components involved in the {\sl INDIGO--DataCloud}
solutions and use cases validation, by deploying and executing the applications
with the last stable version of the software in environments as similar as
possible to production ones. A map of the Pilot Preview
infrastructure is depicted in Figure \ref{fig:fig_pilotpreview}, it shows the
resources providers and the components or services they deployed and supported.

@@

Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are doable, reducing
the impact on production systems. In this regard, a pre--production validation,
through the preview testbeds or resource centers early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production.

Software lifecycle demands as well from a continuous measurement and analysis
process that permits timely reactions to discover and fix defects throughout the
development, release and post--release or maintenance stages. Conscientious metric
tracking and response sits at the core of an appropriate software maturity, as
ultimate reliability can only be achieved by enforcing and empowering the SQA
criteria through each stage of the software lifecycle.

@@

\bibitem{h2020}
\emph{HORIZON 2020}, The EU Framework Programme for Research and Innovation [Online]. Available: https://ec.europa.eu/programmes/horizon2020/

\bibitem{agile-manifesto}
\emph{Manifesto for Agile Software Development} [Online]. Available: http://agilemanifesto.org

@@

\bibitem{cordis:datagrid}
\emph{DataGrid project}, European Community Research and Development
Information Service (CORDIS) [Online]. Available: http://cordis.europa.eu/project/rcn/53665\_en.html

@@

\bibitem{cordis:egee}
\emph{Enabling Grids for E--sciencE (EGEE)} project, European Community
Research and Development Information Service (CORDIS) [Online]. Available:
http://cordis.europa.eu/project/rcn/80149\_en.html

\bibitem{cordis:egee2}
\emph{Enabling Grids for E--sciencE--II (EGEE--II)} project, European Community
Research and Development Information Service (CORDIS) [Online]. Available:
http://cordis.europa.eu/project/rcn/99189\_en.html

\bibitem{cordis:egee3}
\emph{Enabling Grids for E--sciencE--III (EGEE--III)} project, European Community
Research and Development Information Service (CORDIS) [Online]. Available:
http://cordis.europa.eu/project/rcn/87264\_en.html

\bibitem{egee:acceptance-criteria}
\emph{Definition and Documentation of the Revised Software Life--Cycle Process},
Milestone MSA3.4.2, 2010, EGEE--III project [Online]. Available: https://edms.cern.ch/ui/file/1062487/2/EGEE--III--MSA3.4.2-1062487-v1\_4.pdf

@@

\bibitem{cordis:etics}
\emph{E--Infrastructure for Testing, Integration and Configuration of Software
(ETICS)} project, European Community Research and Development Information
Service (CORDIS) [Online]. Available: http://cordis.europa.eu/project/rcn/80138\_en.html

\bibitem{cordis:etics2}
\emph{E--Infrastructure for Testing, Integration and Configuration of Software -
Phase 2 (ETICS 2)} project, European Community Research and Development
Information Service (CORDIS), [Online]. Available: http://cordis.europa.eu/project/rcn/86604\_en.html

@@

\bibitem{cordis:emi}
\emph{European Middleware Initiative (EMI)} project, European Community
Research and Development Information Service (CORDIS) [Online]. Available:
http://cordis.europa.eu/project/rcn/95311\_en.html

\bibitem{iso-9126}
\emph{ISO/IEC 9126 Software Engineering - Product Quality}, International
Organization for Standarization [Online]. Available: https://www.iso.org/standard/22749.html

\bibitem{emi-metrics}
\emph{EMI Metrics Specification} [Online]. Available: https://goo.gl/CCtY7x

\bibitem{emi-quality-model}
\emph{EMI Quality Model} [Online]. Available: https://goo.gl/LdS6fL

\bibitem{cordis:egi-inspire}
\emph{European Grid Initiative: Integrated Sustainable Pan--European
Infrastructure for Researchers in Europe (EGI--InSPIRE)} project, European
Community Research and Development Information Service (CORDIS) [Online]. Available:
http://cordis.europa.eu/project/rcn/95923\_en.html

\bibitem{egi-ops}
\emph{EGI Operations Portal} [Online]. Available:
https://operations-portal.egi.eu/

\bibitem{ggus}
\emph{GGUS} EGI Helpdesk [Online]. Available:
https://ggus.eu/

\bibitem{gocdb}
\emph{GOCDB} EGI Grid Configuration Repository [Online]. Available:
https://ggus.eu/

\bibitem{egi-qc}
\emph{EGI Quality Criteria} [Online]. Available:
https://egi-qc.github.io/

\bibitem{github}
\emph{GITHUB} [Online]. Available:
https://github.com/

\bibitem{jenkins}
\emph{Jenkins} Continuous Integration System [Online]. Available:
https://jenkins-ci.org/

\bibitem{egi-qc}
\emph{EGI Quality Criteria} [Online]. Available: http://egi-qc.github.io/

@@

\bibitem{cordis:egi-engage}
\emph{Engaging the EGI Community towards an Open Science Commons (EGI--ENGAGE)}
project, European Community Research and Development Information Service
(CORDIS), [Online]. Available:  http://cordis.europa.eu/project/rcn/194937\_en.html

\bibitem{cordis:indigo}
\emph{INtegrating Distributed data Infrastructures for Global ExplOitation
(INDIGO--DataCloud)} project, European Community Research and Development
Information Service (CORDIS) [Online]. Available:
http://cordis.europa.eu/project/rcn/194882\_en.html

>>>>

\IEEEPARstart{D}uring the last 15 years the provision of an European e--Infrastructure,
supporting unified access to large--scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan--European research collaborations. One such example is the Worldwide
Large Hadron Collider (LHC) Computing Grid infrastructure, driven
by the particle physics scientific community, built to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
% Comment from reviewer
%(Update: I see this is discussed later... so it would be good just to signpost at the start that a) the focus is really on "grids", and b) that specific examples will be given.
LHC particle accelerator. A significant number of innovation projects and
technology transfer activities -- from the European Commission (EC)-funded 5th, 6th and 7th Framework Programmes,
and recently from the Horizon 2020 programme \cite{h2020} -- have as their main goal to
develop software solutions that provide a seamless and transparent access of researchers to distributed
computing and data resources.

% Comment
%%top of second column:  think about saying this more gently
Major developments were carried out throughout the years focused on providing new
functionalities, partially neglecting the adoption of software engineering methodologies, in the early days.
The released software was less tested, resulting in a high rate of
rollbacks and patches applied to production systems. Thereby, the stability and
reliability of the e--Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.

@@

% Comment
%P1 C2 L18: Claiming that stability and reliability was regularly compromised is quite a bold statement -- it needs at least some citations, and preferably some explanation of what forms this compromise took for a few prominent cases.
Analysing the EC--funded software projects over the last decade,
one can observe a continuous increase in the prominence and robustness of software
testing and validation procedures. The
% Comments
%P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and
operating--system--level virtualization known as containerization) motivated this trend. All of which
complemented by the emergence of automation and event--response tools.

@@

\begin{table*}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Features on software reliability in EC--projects}
\label{tab:feat}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{adjustbox}{max width=0.5\textwidth}
\begin{tabular}{llllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
\hline
\hline
Architecture Task Force            &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling             &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling              &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies                &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection             &       &       &       &$\surd$&       &$\surd$\\
Build/Testing Management Procedure &$\surd$&$\surd$&$\surd$&$\surd$&       &$\surd$\\
Software Product Metrics           &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition        &       &       &$\surd$&$\surd$&$\surd$&$\surd$\\
%Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
Automatic Certification            &       &       &$\surd$&       &$\surd$&$\surd$\\
Auto--generated Documentation      &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD) &       &       &       &       &       &$\surd$\\
\hline
\hline
\end{tabular}
%\end{adjustbox}
\end{table*}

\subsection{Enabling Grids for E--sciencE}
% Comment
%%P2 C2 L24: Typographical: It would be good to merge the three citations into one bracket/\cite
The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee,cordis:egee2,cordis:egee3} projects brought together
scientists and engineers from over 240 institutions in 45 countries aiming to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization (outside Europe) of the project, embracing worldwide research
institutions and user communities. The software to sustain the increasing
requirements coming from the diverse scientific communities needed developing a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15 thousand researchers and deployed in
over 250 institutions.

@@

The {\sl INDIGO--DataCloud} (INtegrating Distributed data Infrastructures for Global
explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this paper. At the time of writing,
the project is succeeding in providing solutions to address the existing gaps in
Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructure administrators and scientific communities to exploit
Cloud computing benefits.

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper, it
leveraged the lessons learned from the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, management of new technologies to put those methodologies
into practice, compilation and analysis of an appropriate set of metrics to measure the
% Comment
%%L26-29 Missing verb
%%(not a sentence—no verb) 
quality of the implemented solutions. All the above with the additional complexity of being framed into
large collaborative projects with an extensive list of partners.

In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the
% Comment
%%P3 C2 L29: This introductory paragraph goes into a bit too much jargony detail, I think, especially given that it is followed by in-depth coverage of each "pillar". I suggest a clearer identification of each pillar, maybe via bullets or definition paragraphs, avoiding or with general-audience definition of jargon like DevOps, delivery mechanisms, two-factor validation, pilot testbeds, and early adoption. I appreciate this may all be highly meaningful to someone steeped in this operational area, but to an interested outsider it reads much like management jargon -- particuarly the final sentence: is this a long-winded way of saying "used during the commissioning period"?
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition} provides guidelines to develop quality
software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing prompt reactions to detected malfunctions throughout the
process. A \textit{bottom--up adoption of DevOps principles} promotes automation to enable
the foregoing change--based development process and set up advanced delivery mechanisms.
The last key pillar is the \textit{two--factor validation} carried out over the project's pilot testbeds and
e--Infrastructure resource provider's early--adoption.

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=\textwidth]{images/fig2.png}
%\includegraphics[width=0.8\textwidth]{images/fig2a.pdf}
\caption{Comparison of the unit testing coverage values for the {\sl INDIGO--DataCloud} software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases.}
\label{fig:fig_unittest}
\end{figure*}

@@

\begin{itemize}
\item \textit{Code style}.
The ultimate goal for code style assessment is to improve the readability and reusability of the source code produced under the scope of the
project. Code style standards are enforced for every software component in the project stack. Community
most--adopted or \textit{de--facto standards} are initially recommended, however the development teams eventually
choose the set of guidelines to comply with. Fig.~\ref{fig:fig_codestyle} highlights the code style standards and
their popularity
in the {\sl INDIGO--DataCloud} software stack.

@@

%% Comment from reviewer
%% These percentages appear as vague approximations by eye, please add the precise numbers rounded.

 \textit{Unit testing} completes the source code testing evaluation by
focusing on the code's internal design. The SQA criteria set a recommended threshold of
70\% coverage for unit testing. In Fig.~\ref{fig:fig_unittest}, unit testing coverage is compared
for each software component between the two major {\sl INDIGO--DataCloud} releases, denoting an incremental
trend over the course of the project. It can be seen that 50\% of the components were over the
70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.

@@

\item \textit{Code review}.
% Comment
%%Represents …   (not a sentence) 
Represents the last step in the change management pipeline, once the candidate change has
successfully passed through the testing methods described previously. It implies the human--based
revision of the proposed change to discuss its adequacy in terms of e.g. scope, objective fulfilment,
and documentation completeness. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
important when the SQA testing requirements are fully automated. Secure code reviews are also
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

\item \textit{Automated Deployment}.
To lower the barriers of software adoption, the SQA criteria require the automated
deployment of the products delivered as part of the catalogue. Automation in this context is
tackled using a configuration management tool. These tools allow to have an
\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation
-- self--documenting code -- as they sequentially guide the component's deployment process in multiple
platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \cite{indigo-galaxy}. Fig.~\ref{fig:fig_confman}
shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release and in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\subsection{DevOps practices}
\label{sec:devops}
% Comment
%P5 C2 L21: Define DevOps!
The DevOps methods emphasize the SQA techniques to avoid infrastructure
disruption whenever new developments are deployed into production systems. The
following DevOps approaches have been progressively adopted throughout the project's
lifetime, being one of its major outcomes in terms of software reliability.

@@

% Comment
%%not a sentence, again
Jenkins \cite{jenkins}, the event--response CI system, and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs.
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline, report their exit status back to GitHub preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects increases. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected in this phase.

Fig.~\ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the
{\sl INDIGO--DataCloud}'s first and second releases, of the required test types builds since the
implementation of the CI infrastructure. Automated functional testing coverage were not
available for all the software stack, thus the associated number of builds are fewer.

\subsubsection{Continuous Delivery}
% Comment from reviewer
%Please add a reference.
As DevOps suggests, frequent releases positively affect the reliability of the software as
they involve fewer changes and usually are more correlated. The software updates of
{\sl INDIGO--DataCloud} products, that has been taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI--validated
software components are automatically delivered in online repositories without the need of
any human interaction but ultimately requiring expert supervision to validate the results.

@@

The \textit{Docker container pipeline} uses the automated deployment solution to
actually install and configure the software component in the container image. Once this is
done the recently created image is uploaded to the production DockerHub repository, tagged
with the label corresponding to the current {\sl INDIGO--DataCloud} release.
Fig.~\ref{fig:fig_CD} shows the complete workflow of the Docker container CD implementation,
automatically triggered by a change in the source code.

@@

Two applications from "Definition of Support to Research Communities" work package,
DisVis \cite{disvis} and PowerFit \cite{powerfit}, were
integrated into a similar CI/CD pipeline described in section \ref{sec:devops}.
Fig.~\ref{fig:fig_disvis} shows the pipeline for the Disvis application.

@@

Two pilot infrastructures are at the disposal of developers and scientific
communities involved in the project. The aim of these testbeds is to test the
level of integration between the components involved in the {\sl INDIGO--DataCloud}
solutions and use cases validation, by deploying and executing the applications
with the last stable version of the software in environments as similar as
possible to production ones. A map of the Pilot Preview
infrastructure is depicted in Fig.~\ref{fig:fig_pilotpreview}, it shows the
resources providers and the components or services they deployed and supported.

@@

% Comment
%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies? Is there some quantitative evidence of uptime, or bug reports per user, or similar? Or some evidence of speed of defect detection & fixing, which you cite as greatly improved? I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are doable, reducing
the impact on production systems. In this regard, a pre--production validation,
through the preview testbeds or resource centers early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production.

% Comment
%%L42-44 not clear, please rephrase
Software lifecycle demands as well from a continuous measurement and analysis
process that permits timely reactions to discover and fix defects throughout the
development, release and post--release or maintenance stages. Conscientious metric
tracking and response sits at the core of an appropriate software maturity, as
ultimate reliability can only be achieved by enforcing and empowering the SQA
criteria through each stage of the software lifecycle.

@@

\bibitem{h2020}
\emph{HORIZON 2020}, The EU Framework Programme for Research and Innovation [Online]. Available: https://ec.europa.eu/programmes/horizon2020/ .

\bibitem{agile-manifesto}
\emph{Manifesto for Agile Software Development} [Online]. Available: http://agilemanifesto.org .

@@

\bibitem{cordis:datagrid}
\emph{DataGrid project}, European Community Research and Development
Information Service (CORDIS) [Online]. Available: http://cordis.europa.eu/project/rcn/53665\_en.html .

@@

\bibitem{cordis:egee}
\emph{Enabling Grids for E--sciencE (EGEE)} project, European Community
Research and Development Information Service (CORDIS) [Online]. Available:
http://cordis.europa.eu/project/rcn/80149\_en.html .

\bibitem{cordis:egee2}
\emph{Enabling Grids for E--sciencE--II (EGEE--II)} project, European Community
Research and Development Information Service (CORDIS) [Online]. Available:
http://cordis.europa.eu/project/rcn/99189\_en.html .

\bibitem{cordis:egee3}
\emph{Enabling Grids for E--sciencE--III (EGEE--III)} project, European Community
Research and Development Information Service (CORDIS) [Online]. Available:
http://cordis.europa.eu/project/rcn/87264\_en.html .

\bibitem{egee:acceptance-criteria}
\emph{Definition and Documentation of the Revised Software Life--Cycle Process},
Milestone MSA3.4.2, 2010, EGEE--III project [Online]. Available: https://edms.cern.ch/ui/file/1062487/2/EGEE--III--MSA3.4.2-1062487-v1\_4.pdf .

@@

\bibitem{cordis:etics}
\emph{E--Infrastructure for Testing, Integration and Configuration of Software
(ETICS)} project, European Community Research and Development Information
Service (CORDIS) [Online]. Available: http://cordis.europa.eu/project/rcn/80138\_en.html .

\bibitem{cordis:etics2}
\emph{E--Infrastructure for Testing, Integration and Configuration of Software -
Phase 2 (ETICS 2)} project, European Community Research and Development
Information Service (CORDIS), [Online]. Available: http://cordis.europa.eu/project/rcn/86604\_en.html .

@@

\bibitem{cordis:emi}
\emph{European Middleware Initiative (EMI)} project, European Community
Research and Development Information Service (CORDIS) [Online]. Available:
http://cordis.europa.eu/project/rcn/95311\_en.html .

\bibitem{iso-9126}
\emph{ISO/IEC 9126 Software Engineering - Product Quality}, International
Organization for Standarization [Online]. Available: https://www.iso.org/standard/22749.html .

\bibitem{emi-metrics}
\emph{EMI Metrics Specification} [Online]. Available: https://goo.gl/CCtY7x .

\bibitem{emi-quality-model}
\emph{EMI Quality Model} [Online]. Available: https://goo.gl/LdS6fL .

\bibitem{cordis:egi-inspire}
\emph{European Grid Initiative: Integrated Sustainable Pan--European
Infrastructure for Researchers in Europe (EGI--InSPIRE)} project, European
Community Research and Development Information Service (CORDIS) [Online]. Available:
http://cordis.europa.eu/project/rcn/95923\_en.html .

\bibitem{egi-ops}
\emph{EGI Operations Portal} [Online]. Available:
https://operations-portal.egi.eu/ .

\bibitem{ggus}
\emph{GGUS} EGI Helpdesk [Online]. Available:
https://ggus.eu/ .

\bibitem{gocdb}
\emph{GOCDB} EGI Grid Configuration Repository [Online]. Available:
https://goc.egi.eu/ .

\bibitem{egi-qc}
\emph{EGI Quality Criteria} [Online]. Available:
https://egi-qc.github.io/ .

\bibitem{github}
\emph{GITHUB} [Online]. Available:
https://github.com/ .

\bibitem{jenkins}
\emph{Jenkins} Continuous Integration System [Online]. Available:
https://jenkins-ci.org/ .

\bibitem{egi-qc}
\emph{EGI Quality Criteria} [Online]. Available: http://egi-qc.github.io/ .

@@

\bibitem{cordis:egi-engage}
\emph{Engaging the EGI Community towards an Open Science Commons (EGI--ENGAGE)}
project, European Community Research and Development Information Service
(CORDIS), [Online]. Available:  http://cordis.europa.eu/project/rcn/194937\_en.html .

\bibitem{cordis:indigo}
\emph{INtegrating Distributed data Infrastructures for Global ExplOitation
(INDIGO--DataCloud)} project, European Community Research and Development
Information Service (CORDIS) [Online]. Available:
http://cordis.europa.eu/project/rcn/194882\_en.html .

\bibitem{indigo-dockerhub}
\emph{INDIGO--DataCloud DockerHub repository} [Online]. Available:
https://hub.docker.com/u/indigodatacloud .
